A Comparison of Neural Network Models for Water Quality Forecasting

### 2024-06-13 manuscript/manuscript.Rmd
Additions on disc

Over the last decade, machine learning has gained significant prominence after a spate of high profile successes based on neural network models, some of these successes include beating the master champion of Go, accurately predicting protein structures and the advent of large language models like ChatGPT \cite{alphago, alphafold, chatgpt}. During this time, researchers in computer science have developed new neural network models for time series forecasting with promises of significant accuracy improvements over previous methods \cite{makridakis}. Simultaneously during the last 10 years, there has been a growing focus in limnology towards using neural networks for water quality forecasting, but this focus has primarily centered on older neural network architectures and there has been little exploration of newer methods. This study aims to address this research gap by performing a comprehensive comparison of neural network models, including models taken from more recent research in computer science.

Instead of identifying that there is some new neural network model that performs exceptionally well, this study has affirmed the well-established forecasting axiom that ensemble forecasts are often more accurate and robust than forecasts from single models. We found that the Naive Ensemble model which aggregated the forecasts from the neural network models was the best performing model overall, evincing all the other models on the target variables of dissolved oxygen and water temperature, while also performing well for chlorophyll-a (attaining the second and third best CRPS and RMSE, respectively). This result affirms recent work from \cite{freya} that established multi-model ensembles as advantageous over individual models for water temperature forecasting.

Yet, as is often the case when judging whether one method is superior to another, we have found that subtle changes to our criteria may have produced radically different results. For instance, it is possible that if we changed the forecast horizon to 5 days, then the AutoTheta model may have been the best performing model as the AutoTheta model outperformed all the other models in short time horizons (see Figure \ref{}). But, it is important to qualify that changing our criteria to consider shorter term forecasts may result in the neural network models learning different behavior during training. How these models perform with different forecast horizons warrants future research.

The neural network models struggled to forecast Chlorophyll-A, a result that is not surprising given that forecasting algae blooms has been well established as a challenging prediction problem in limnology \cite{}. The neural network models displayed a tendency to take a conservative strategy, regularly predicting that the chlorophyll-a concentration would remain in a non-bloom state. It is possible that the neural network models settled on this conservative behavior as they were not provided enough information to predict blooms -- the neural networks were not provided with some of the typical covariates used in process-based chlorophyll-a models like Nitrogren and Phosphorous concentrations as well as measures of photosynthetic active radiation. And, it is likely that this conservative strategy is an artifact of the length of the forecast horizon. Blooms are stochastic events, characterized by rapid fluctuations in chlorophyll-a. Predicting the exact time of such an event in a long term horizon will be more difficult than predicting the onset of a bloom in the near-term \cite{}.

We hope that this comparative study will encourage more research that investigates neural networks for water quality forecasting. In place of the one research question that we attempted to answer, "How do neural network models perform for water quality forecasting?", many new questions and ideas have arose. Outside of exploring how the performance of neural networks will vary with forecast horizon, which was mentionned previously, we think that some other promising directions for exploration include cross-learning and hybrid models.

Provided that there are large limnological datasets that currently exist, and there will be an increasing amount of limnological data coming online, neural network models offer significant advantages over process-based and statistical models as they will be able to learn directly from the data. This data-driven approach is advantageous in a few ways: as NN models can learn directly from the data, these models will not be limited by the assumptions inherent to process-based and statistical models. Moreover, these models should improve as

>>>>

Over the last decade, machine learning has gained significant prominence after a spate of high profile successes based on neural network models, some of these successes include beating the master champion of Go, accurately predicting protein structures and the advent of large language models like ChatGPT \cite{alphago, alphafold, chatgpt}. During this time, researchers in computer science have developed new neural network models for time series forecasting with promises of significant accuracy improvements over previous methods \cite{makridakis}. Simultaneously during the last 10 years, there has been a growing focus in limnology towards using neural networks for water quality forecasting, but this focus has primarily centered on older neural network architectures and there has been little exploration of newer methods. This study aims to address this research gap by performing a comprehensive comparison of neural network models, including models taken from more recent developments in computer science.

Instead of identifying that there is some new neural network model that performs exceptionally well, this study has affirmed the well-established forecasting axiom that ensemble forecasts are often more accurate and robust than forecasts from single models. We found that an ensemble model which aggregates the forecasts from the neural network models was the best performing model overall, evincing all the other models on the target variables of dissolved oxygen and water temperature, while also performing well for chlorophyll-a, attaining the second and third best CRPS and RMSE, respectively. This result affirms recent work from \cite{freya} which established that multi-model ensembles offer advantages over individual models for water temperature forecasting.

Yet, as is often the case when judging whether one method is superior to another, we have found that subtle changes to our criteria may have produced radically different results. For instance, if did not change any of the model training configurations but cut off the forecast horizon to 5 days, then the AutoTheta model would have been the best performing model since the AutoTheta model outperformed all the other models in short time horizons (see Figure \ref{}). But, it is important to qualify that if we changed the model training configurations to consider shorter forecasts horizons, this could result in the neural network models learning different behavior during training. How neural networks models perform with different forecast horizons warrants future research.

The neural network models struggled to forecast Chlorophyll-A, a result that is not surprising given that forecasting algae blooms is well established as a challenging prediction problem in limnology \cite{}. With Cholorophyll-A (chla), the neural network models displayed a tendency to take a conservative strategy, regularly predicting that the chla concentration would remain in a non-bloom state. It is possible that the neural network models settled on this conservative behavior because we did provide enough information to predict blooms; the neural networks were not provided with some of the typical covariates used in process-based models like Nitrogen and Phosphorous concentrations, and photosynthetic active radiation. It is also likely that this conservative strategy is an artifact of the length of the forecast horizon. Blooms are stochastic events, characterized by rapid fluctuations in chlorophyll-a. Predicting the exact time of such an event in a long term horizon will be more difficult than predicting the onset of a bloom in the near-term \cite{}. It seems reasonable that the neural networks would take a conservative approach to forecasting chla as making the wrong prediction for an algae bloom will be costly with respect to the model's objective.

In place of the one research question that we set out to answer, "How do neural network models perform for water quality forecasting?", many new questions and ideas have arose from this project. Outside of exploring how the performance of neural networks will vary with forecast horizon, we think that some other promising directions for exploration include cross-learning and hybrid models. Throughout this study, we have trained models individually for each target time series; cross-learning presents a different approach where we would train one model across all the time series available for a given target variable. Through cross-learning, the model might be able to learn a pattern from one time series that will improve forecasts on a different time series; but, if the model is trained on a single time series, then the model may never learn this pattern. We also reckon that hybrid models may have superior forecasting performance to neural network or statistical models individually. We observed that AutoTheta were the best performing models early in the forecast horizon, and neural networks performed best later in the forecast horizon. This motivates the exploration of a hybrid model that employs AutoTheta at the beginning of the forecast horizon and neural network models later in the forecast horizon.

It is an exciting time to study water quality forecasting. There are large, publicly accessible water quality data sets that exist now, and there will be an increasing amount of limnological data in the future as more sensors will be deployed \cite{}. Neural network models offer promising advantages over alternative methods for limnological forecasting as these methods can learn complex patterns without being restrained by a need for domain knowledge. Furthermore, their performance has the potential to improve as more water quality data becomes available. Yet, they also present a range of problems like a lack of interpretability and generalizability that could restrain them from being useful decision support tools for the safety critical problem of water resource management. This study does not attempt to provide a definitive answer as to whether neural networks will be the best method for water quality forecasting as answering this questions will take numerous studies to reach a consensus on. Instead, we hope that this study will generate ideas and a collective momentum towards improving our general ability to forecast and manage water quality.
